Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents

### 2025-05-21 doc/cicm25-paper/paper.tex
more

\paragraph{Overview \& Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterweards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

\paragraph{Acknowledgments}
The work reported in this article was conducted as part of the VoLL-KI project (see
\url{https://voll-ki.de}) funded by the German Research/Education Ministry under grant
16DHBKI089.

@@

In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
object.  Verbalizations can vary between between subdomains and communities of
practice. In our example, the verbalization ``terminal [state]'' was annotated with the
symbol with name \lstinline|goal state|. It comes from the module with name
\lstinline|search-problem| in the namespace \lstinline|smglom/search/mod| imported via the
\lstinline|\importmodule| macro in line 1. As the symbol name \lstinline|goal state| is
unique in the imports here it suffices as a relative URI in the \lstinline|\sr|
annotation. If the symbol name and the desired verbalization coincide (as they often do in
English), the short-hand macro \lstinline|\sn| can be used (e.g. as
\lstinline|\sn{utility}| for ``utility'' in \Cref{lst:losa}).

@@

To annotate the word ``terminal'' in the document, the author has to be aware that it is a
technical term, that the module from \Cref{fig:state-space} exists, know the symbol URI,
and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
directive if it is not already (recursively) implied and possibly removing directives that
become redundant by the new one.  At ca 150,000 words in a semester's worth of lecture
notes and ca. 10--15\% of technical terms a rather daunting task.

@@

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model -- a
highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors.\ednote{MK: this shared domain model
  could be thought of as the informal content commons; hmmm, but maybe the literature is
  the content commons} Consequently a large part of the authoring problem lies in
predicting the state and extent of the domain model of the reader and creating text that
bridges the gap between the reader's domain model and the payload content of the article
-- the knowledge the article or textbook intends to impart -- ideally using the technical
vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
if we disregard LLM-based writing/formulation support for now.

\paragraph{Formal Authoring}
In fully formal authoring, e.g. programming, the content commons is given formally in
the form of the libraries of the respective programming language, correspondingly the
authored content -- a program -- is also formal and intended for processing by a computer,
for which the formality is a prerequisite. Formal authoring is usually supported by an
integrated development environment (IDE) that makes use of the fully formal content and
offers services like semantic highlighting, tab completion, jumping to variable
declarations, function definitions, and interface definitions, as well as refactoring
support. All of these rely on a syntactic/semantic analysis of the formal content. The
situation is very similar for authoring of formal knowledge, e.g. in program verification
or the formalization of mathematics.

@@

\begin{newpart}{MK: please re-read}
  A distinctive feature of semantic authoring that is not apparent at the first glance is
  that the added investment in annotation moves the sweet spot in the re-use/customize
  trade-off towards sharing and interoperability. In our experience sufficiently annotated
  documents can be automatically customized if the semantics has multiple
  notations/verbalizations/presentations. Consider for instance course materials on
  complex arithmetic, which become re-usable across the electrical engineering/mathematics
  border if the imaginary unit in formulae is represented e.g. by a semantic macro
  \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths)
  and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
  creating reusable/shareable document fragments organized in central or federated
  repositories and hubs. This is quite similar to the formal authoring system, where we
  see a tendency towards massively shared math libraries in formal systems and package
  hubs for programming libraries. As a consequence, semantic authoring support systems
  should facilitate the logistics of using and curating shared libraries.
\end{newpart}

@@

Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
There also is prototypical support for authoring \ALeA content in Microsoft Word with
the \WOIDE plugin \cite{KohKoh:woide24}, which allows authors to use Word's spell
checker and grammar checker.\ednote{MK: Mention CPoint and SMD as well? Probably yes.}

@@

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline[language={}]|\emph|
or \lstinline[language={}]|\textbf|), and then steps through them similarly to \snify.  For each
potential definiendum, the user can then use fuzzy search to find the corresponding symbol
in the domain model.  This populates the catalog with verbalizations relevant to the
document, making subsequent \snify runs more effective, especially for documents in
languages that are not well covered by the domain model (i.e.\ not English).

@@

The general situation is that on the one hand there is a constantly growing, shared
\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
other hand, authors are working on
\begin{compactenum}[\em i\rm)]
\item a set of local working copies of the archives -- with the intention of publishing
  them on MathHub.info -- and their dependencies (they need to be local since pdflatex
  cannot deal with remote files) and
\item possibly a set of private archives (e.g.\ papers under development or exam
  problems/solutions).
\end{compactenum}
This situation is familiar from software development and also from e.g. the \LaTeX{}
ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
from both.

\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node[draw] (sn) at (0,4.5) {\snify};
    \node[draw] (id) at (2.5,5) {IDE};
    \node[draw] (wo) at (5,4.5) {\WOIDE};
    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
    \node (f) at (-.5,.9) {\flams};
    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[includeleft] (pl) to[bend right=10] (wo);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
\end{wrapfigure}
To cope with this situation, we are working on extending the brand-new \flams system
\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
from the flexiformal content commons would be just a simple SPARQL query and some postprocessing
of the result (de-duplication, stemming, and pairing) away. Note that \flams works
directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}.

@@

In this situation, the main functionality of \snify would be to maintain the verbalization cache
and answer front-end queries for verbalizations efficiently so that we can power various
front-ends from that. These include
\begin{compactenum}[\em i\rm)]
\item The traditional text/terminal-based \snify UI for bulk annotation,
\item the \sTeX IDEs in \textsf{VSCode} and \textsf{emacs} that could supply annotation
  support during editing, and 
\item \WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
    CPoint and semantic wikis.}
\end{compactenum}

@@

The bulk annotation workflow for existing informal text
is also relevant when creating machine learning datasets,
e.g.\ for named entity recognition.
There are many (often browser-based) tools for this task;
a prominent example is \textsf{brat}~\cite{brat:on}.

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs

>>>>

\paragraph{Overview \& Contribution}
In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
Using \sTeX/\ALeA as an exemplary case study,
we will provide an overview of the authoring support (\Cref{sec:tools})
and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\Cref{sec:snify}).
Afterwards, we discuss on-going work to integrate \snify better
into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).

\paragraph{Acknowledgments}
The work reported in this article was conducted as part of the VoLL-KI project (see
\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMBF) under
grant 16DHBKI089.

@@

In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
object.  Verbalizations can vary between between subdomains and communities of
practice. In our example, the verbalization -- i.e. the natural language phrase used to
express the symbol in English -- ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in
\Cref{lst:losa}).

@@

To annotate the word ``terminal'' in the document, the author has to 
\begin{inparaenum}[\em i\rm)]
\item be aware that it is a
technical term, 
\item that the module from \Cref{fig:state-space} exists,
\item know the symbol URI, 
\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
  directive if it is not already (recursively) implied and possibly removing directives
  that become redundant by the new one.
\end{inparaenum}
At ca 150,000 words in a semester's worth of lecture notes and ca. 10--15\% of technical
terms a rather daunting task.

@@

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
a highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors. Consequently a large part of the
authoring problem lies in predicting the state and extent of the domain model of the
reader and creating text that bridges the gap between the reader's domain model and the
payload content of the article -- the knowledge the article or textbook intends to impart
-- ideally using the technical vocabulary the reader is familiar with to reduce this
gap. Tool support for traditional authoring is minimal and usually restricted to
spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
support for now.

\paragraph{Formal Authoring}
In programming, the content commons is given formally in the form of the libraries of the
respective programming language, correspondingly the authored content -- a program -- is
also formal and intended for processing by a computer, for which the formality is a
prerequisite. Formal authoring is usually supported by an integrated development
environment (IDE) that makes use of the fully formal content and offers services like
semantic highlighting, tab completion, jumping to variable declarations, function
definitions, and interface definitions, as well as refactoring support. All of these rely
on a syntactic/semantic analysis of the formal content. The situation is very similar for
authoring of formal knowledge, e.g. in program verification or the formalization of
mathematics.

@@

A distinctive feature of semantic authoring that is not apparent at the first glance is
that the added investment in annotation moves the sweet spot in the re-use/customize
trade-off towards sharing and interoperability. In our experience sufficiently annotated
documents can be automatically customized if the semantics has multiple
notations/verbalizations/presentations. Consider for instance course materials on complex
arithmetic, which become re-usable across the electrical engineering/mathematics border if
the imaginary unit in formulae is represented e.g. by a semantic macro
\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
reusable/shareable document fragments organized in central or federated repositories and
hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
massively shared math libraries in formal systems and package/library hubs for programming
libraries. As a consequence, semantic authoring support systems should facilitate the
logistics of using and curating shared libraries.

@@

Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
There also is prototypical support for authoring \ALeA content in Microsoft Word with
the \WOIDE plugin \cite{KohKoh:woide24}, which allows authors to use Word's spell
checker and grammar checker.

@@

\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to consider different \LaTeX\ macros to identify definienda (e.g.\
\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
through them as \snify does.  For each potential definiendum, the user can then use fuzzy
search to find the corresponding symbol in the domain model.  This populates the catalog
with verbalizations relevant to the document, making subsequent \snify runs more
effective, especially for documents in languages that are not well covered by the domain
model (i.e.\ not English).

@@

The general situation is that on the one hand there is a constantly growing, shared
\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
other hand, authors are working on
\begin{compactenum}[\em i\rm)]
\item a set of local working copies of the archives -- with the intention of publishing
  them on MathHub.info -- and their dependencies (they need to be local since
  \textsf{pdflatex} cannot deal with remote files) and
\item possibly a set of private archives (e.g.\ papers under development or exam
  problems/solutions).
\end{compactenum}
This situation is familiar from software development and also from e.g. the \LaTeX{}
ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
from both.

\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node[draw] (sn) at (0,4.5) {\snify};
    \node[draw] (id) at (2.5,5) {IDE};
    \node[draw] (wo) at (5,4.5) {\WOIDE};
    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
    \node (f) at (-.5,.9) {\flams};
    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[includeleft] (pl) to[bend right=10] (wo);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
\end{wrapfigure}
To cope with this situation, we are working on extending the brand-new \flams system
\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
from the flexiformal content commons would be just a simple SPARQL query and some
postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
more systems that could profit from this: The CPoint system that allows annotating MS
PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
FTML instead of plain HTML.

@@

In this situation, the main functionality of \snify would be to maintain the verbalization cache
and answer front-end queries for verbalizations efficiently so that we can power various
front-ends from that. These include
\begin{compactenum}[\em i\rm)]
\item The traditional text/terminal-based \snify UI for bulk annotation,
\item the \sTeX IDEs in \textsf{VSCode} and \textsf{emacs} that could supply annotation
  support during editing, and 
\item \WOIDE, which can be seen as a semantic authoring IDE based on MS Word.
\end{compactenum}

@@

The bulk annotation workflow for existing informal text is also relevant when creating
machine learning datasets, e.g.\ for named entity recognition.  There are many (often
browser-based) tools for this task; a prominent example is
\textsf{brat}~\cite{brat:on}.\ednote{MK@FS: discuss semantic wikis}

In all cases, the setting, solution, and resulting workflows differ from the one presented
in this paper. The WissKI system \cite{goerz2010adaptation} has a rudimentary text edition
facility that allows annotating document with ontology -- mostly ABox -- references; this
is the nearest match to our \snify system.

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF
